### 2020-07-17 paper.tex
tweaks to section 4

As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}a, which defines a Gaussian mixture model with an unknown number of components:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function p(n::Int)
  k ~ poisson_plus_one(1)
  means = [
    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
  vars = [
    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
  weights ~ dirichlet([2.0 for j in 1:k])
  for i in 1:n
    {(:x, i)} ~ mixture_of_normals(
        weights, means, vars)
  end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
a distinct distribution over dictionaries.

@@

Given two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfying a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}). (In this case, the program $\mathcal{Q}$ 
accepts a trace $x$ of $\mathcal{P}$ as input, and the distribution for $q_x$ is given by $\mathcal{Q}_x$.)
The acceptance probability of involutive MCMC includes the following ratio of densities:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
\end{equation}

This ratio can be computed term-wise using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$.

However, typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC). Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses. Let $\tilde{p}$ denote $\mathcal{P}$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}. Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{\mathcal{L}(b)}$, where $\mathcal{L}(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the marginal likelihood of $b$, and does not depend on $x$. Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$. If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio above, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator.

Thus, this ratio can be computed term-by-term using \textsc{score} (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$. The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.

@@

>>>>

As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}b, which defines a Gaussian mixture model with an unknown number of components:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function p(n::Int)
  k ~ poisson_plus_one(1)
  means = [
    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
  vars = [
    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
  weights ~ dirichlet([2.0 for j in 1:k])
  for i in 1:n
    {(:x, i)} ~ mixture_of_normals(
        weights, means, vars)
  end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
a distinct distribution over dictionaries.

@@

If the two probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ satisfy a simple technical requirement (described in Section~\ref{sec:technical-requirement}), we can interpret them as specifying the model density $p$ and auxiliary densities $q_x$ for use with involutive MCMC on dictionaries (Algorithm~\ref{alg:involutive-mcmc}). (In this case, the program $\mathcal{Q}$ 
accepts a trace $x$ of $\mathcal{P}$ as input, and the distribution for $q_x$ is given by $\mathcal{Q}_x$.)

However, typically, we cannot use a probabilistic program $\mathcal{P}$ to represent the target distribution directly: probabilistic programs implement simulators for a distribution, but target densities are typically not tractable to simulate (hence the need for MCMC). Instead, we may wish to sample from a target $p$ that arises from conditioning a probabilistic program $\mathcal{P}$'s distribution over traces on observations of the values at some addresses. Let $\tilde{p}$ denote $\mathcal{P}$'s distribution over traces, and let $b = (B, \textbf{b})$ be a dictionary of observations, as described in Section~\ref{sec:dists-on-dicts}. Then, as shown in that section, the target distribution $p(x) = \tilde{p}(x \mid b) = \frac{\tilde{p}(x \oplus b)}{\mathcal{L}(b)}$, where $\mathcal{L}(b) = \int_{\{m \mid m \cap B = \emptyset\}} p(m \oplus b) \mu_\mathcal{D}(\text{d}m)$ is the marginal likelihood of $b$, and does not depend on $x$. Then $\tilde{p}(x \oplus b) = \mathcal{L}(b)p(x)$. If we use $\tilde{p}(x \oplus b)$ in place of $p(x)$ to compute the ratio of densities in Algorithm~\ref{alg:involutive-mcmc}, we will wind up with the same output, because $\mathcal{L}(b)$ will cancel in the numerator and denominator:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)}
= \frac{\tilde{p}(x' \oplus b) q_{x'}(y')}{\tilde{p}(x \oplus b) q_x(y)}
= \frac{\tilde{p}(x' | b) q_{x'}(y')}{\tilde{p}(x | b) q_x(y)}
\end{equation}

Thus, this ratio can be computed term-by-term using \textsc{score} on the probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$: (Algorithm~\ref{alg:auto-involutive-mcmc}): to compute $q_x(y')$, we run the algorithm directly on the program $\mathcal{Q}_x$, and to compute $p(x)$ and $p(x')$, we actually merge the dictionary $x$ with the observations $b$ and compute $\mathcal{L}(b)p(x) = \tilde{p}(x \oplus b)$ instead, by running the algorithm on $\mathcal{P}$ with trace $x \oplus b$. The density $q_x(y)$ can be computed while $y$ is being sampled, using \textsc{trace-and-score}.

@@

Figure~\ref{fig:mixture}b and Figure~\ref{fig:mixture}c show examples of probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$ respectively, for a split-merge reversible jump move.
